\documentclass[notes]{subfiles}

\begin{document}

\setcounter{section}{11}
\section{Lecture (2/26/25)}

\subsection{Solving Constant Coefficient Inhomogeneous Linear ODEs}
Given an order $n$ inhomogeneous linear ODE and consider a particular solution $f_p$ the driver function $g$. The idea is that the particular solution and driver function should be in the same class of function. Therefore, if $g$ is a polynomial, then $f_p$ will also be a polynomial. Similarly, if $g$ is an exponential, then $f_p$ will also be an exponential.

It is important to remember that the particular solution must be linearly independent from the basis solutions.

\begin{exercise}
    Solve $y'' + 4y' - 2y = 2x^2 - 3x + 6$.
\end{exercise}
\begin{solution}
    First we need to solve the associated homogenous ODE. Therefore consider $y = e^{mx}$ so the characteristic polynomial is
    \begin{align*}
        0 = m^2 + 4m - 2
        &\iff m = \frac{-4 \pm \sqrt{16 + 8}}{2} = -2 \pm \sqrt{6}
    \end{align*}
    Therefore the null solution is $y_n = c_1e^{(-2 - \sqrt{6})x} + c_2e^{(-2 + \sqrt{6})x}$. Since the driver is a polynomial then we guess $y_p = Ax^2 + Bx + C$, therefore
    \[
        y_p' = 2Ax + B \quad \text{and} \quad y_p'' = 2A
    \]
    So we get
    \begin{align*}
        &y_p'' + 4y_p' - 2y_p = 2x^2 - 3x + 6 \\
        \iff& 2A + 4(2Ax + B) - 2(Ax^2 + Bx + C) = 2x^2 - 3x + 6 \\
        \iff& -2Ax^2 + (8A - 2B)x + (2A + 4B - 2C) = 2x^2 - 3x + 6 \\
        \iff& -2A = 2 \iff A = -1 \\
        \iff& -8 - 2B = -3 \iff -2B = 5 \iff B = -\frac{5}{2} \\
        \iff& -2 - 10 - 2C = 6 \iff -12 - 2C = 6 \iff -2C = 18 \iff C = -9 \\
    \end{align*}
    Therefore $y_p = -x^2 - \frac{5}{2}x - 9$ so
    \[
        y = c_1e^{(-2 - \sqrt{6})x} + c_2e^{(-2 + \sqrt{6})x} - x^2 - \frac{5}{2}x - 9
    \]
    solves the ODE.
\end{solution}

We can also consider each part of a particular solution separately due to superposition.

\begin{exercise}
    Solve $y'' - 2y' - 3y = 4x - 5 + 6xe^{2x}$.
\end{exercise}
\begin{solution}
    For the homogeneous equation consider the characteristic polynomial so
    \begin{align*}
        0 = m^2 - 2m - 3 = (m - 3)(m + 1)
        &\iff m = 3, m = -1
    \end{align*}
    Therefore $y_n = c_1e^{-x} + c_2e^{3x}$. Now we let $y_p = y_{p_1} + y_{p_2}$ where
    
    $y_{p_1} = Ax + B$ and $y_{p_2} = (Cx + D)e^{2x}$.

    For $y_{p_1}$
    \[
        y_{p_1}' = A \quad \text{and} \quad y_{p_1}'' = 0
    \]
    Therefore
    \begin{align*}
        &y_{p_1}'' - 2y_{p_1}' - 3y_{p_1} = 4x - 5 \\
        \iff& -2A - 3(Ax + B) = 4x - 5 \\
        \iff& -3Ax + (-2A - 3B) = 4x - 5 \\
        \iff& -3A = 4 \iff A = -\frac{4}{3} \\
        \iff& \frac{8}{3} - 3B = -5 \iff 3B = \frac{8}{3} + 5 = \frac{23}{3} \iff B = \frac{23}{9} \\
        \iff& y_{p_1} = -\frac{4}{3}x + \frac{23}{9}
    \end{align*}
    For $y_{p_2}$
    \[
        y_{p_2}' = Ce^{2x} + 2(Cx + D)e^{2x} = (2Cx + C + 2D)e^{2x}
    \]
    and
    \[
        y_{p_2}'' = 2Ce^{2x} + 2(2Cx + C + 2D)e^{2x} = (4Cx + 4C + 4D)e^{2x}
    \]
    Therefore
    \begin{align*}
        &y_{p_2}'' - 2y_{p_2}' - 3y_{p_2} = 6xe^{2x} \\
        \iff& (4Cx + 4C + 4D)e^{2x} - 2(2Cx + C + 2D)e^{2x} - 3(Cx + D)e^{2x} = 6xe^{2x} \\
        \iff& 4Cx + 4C + 4D - 2(2Cx + C + 2D) - 3(Cx + D) = 6x \\
        \iff& (4C - 4C - 3C)x + (4C + 4D - 2C - 4D - 3D) = 6x \\
        \iff& -3Cx + (2C - 3D) = 6x \\
        \iff& -3C = 6 \iff C = -2 \\
        \iff& -4 - 3D = 0 \iff 3D = -4 \iff D = -\frac{4}{3} \\
        \iff& y_{p_2} = \left( -2x - \frac{4}{3} \right)e^{2x}
    \end{align*}
    Therefore $y_p = -\frac{4}{3}x + \frac{23}{9} + \left( -2x - \frac{4}{3} \right)e^{2x}$ so
    \[
        y = c_1e^{-x} + c_2e^{3x} - \frac{4}{3}x + \frac{23}{9} + \left( -2x - \frac{4}{3} \right)e^{2x}
    \]
    solves the ODE.
\end{solution}

However, consider the situation where a reasonable guess for a particular solution would be linearly dependent with the basis solutions.

\begin{exercise} \label{linearly_dep_part_sol}
    Solve $y'' - 5y' + 4y = 8e^x$.
\end{exercise}
\begin{solution}
    For the homogeneous ODE, consider the characteristic polynomial
    \begin{align*}
        0 = m^2 - 5m + 4 = (m - 4)(m - 1)
        &\iff m = 4, m = 1
    \end{align*}
    Therefore $y_n = c_1e^x + c_2e^{4x}$.
    However, $y_p \neq Ae^x$ since it would be linearly independent with $y_n$. We can do something similar to reduction of order and bump up the exponent so let $y_p = Axe^x$ and consider
    \[
        y_p' = Ae^x + Axe^x = (Ax + A)e^x \quad \text{and} \quad y_p'' = Ae^x + (Ax + A)e^x = (Ax + 2A)e^x
    \]
    Therefore
    \begin{align*}
        &y_p'' - 5y_p' + 4y_p = 8e^x \\
        \iff& (Ax + 2A)e^x - 5(Ax + A)e^x + 4Axe^x = 8e^x \\
        \iff& (Ax + 2A) - 5(Ax + A) + 4Ax = 8 \\
        \iff& (A - 5A + 4A)x + (2A - 5A) = 8 \\
        \iff& -3A = 8 \iff A = -\frac{8}{3} \\
        \iff& y_p = -\frac{8}{3}xe^x
    \end{align*}
    Therefore $y = c_1e^x + c_2e^{4x} - \frac{8}{3}xe^x$ solves the ODE.
\end{solution}

In page $146$ of the textbook there is a table of particular solution guesses.

\subsection{Annihilation Operators}

\begin{definition}[Annihilation Operator]
    A linear differential operator $L$ is an \textsl{annihilation operator} for $g$ iff $Lg = 0$. 
\end{definition}

The idea for these is to find what linear homogeneous differential equation $g$ solves.

\begin{exercise}
    Find annihilation operators for
    \begin{enumerate}[label = (\alph*)]
        \item $g(x) = e^{3x}$
        \item $g(x) = 5x^3 - 2x + 3$
        \item $g(x) = \cos(2x)$
    \end{enumerate}
\end{exercise}
\begin{solution}
    ~

    \begin{enumerate}[label = (\alph*)]
        \item Consider $g'(x) = 3e^{3x}$, then $g'(x) - 3g(x) = 0$ so $D - 3$ is an annihilation operator for $g$.
        \item We can just differentiate this polynomial $4$ times to get $0$ so $D^4$.
        \item Consider $g'(x) = -2\sin(2x)$ and $g''(x) = -4\cos(x)$ so $g''(x) + 4g(x) = 0$. Therefore $D^2 + 4$ is an annihilation operator for $g$.
    \end{enumerate}
\end{solution}

We can now use this to guess particular solutions when we are stuck.

In \cref{linearly_dep_part_sol} consider that if $g(x) = 8e^x$ then the linear differential operator representation of the equation is $D^2 - 5D + 4 = g$. Since $D - 1$ is an annihilation operator for $g$ we can now apply it to both sides of the equation so
\[
    (D - 1)(D^2 - 5D + 4) = (D - 1)g = 0
\]
Now we solve the resulting homogeneous ODE by considering its characteristic polynomial so
\begin{align*}
    &0 = (m - 1)(m^2 - 5m + 4) = (m - 1)(m - 4)(m - 1) = (m - 1)^2(m - 4) \\
    \iff& m = 1, m = 4
\end{align*}
Therefore, $y = c_1e^x + c_2xe^x + c_3e^{4x}$ solves the resulting homogeneous ODE. Notice that we gain a basis solution of $c_2xe^x$. Therefore the particular solution must be in the form $c_2xe^x$ which is exactly what we saw in \cref{linearly_dep_part_sol}. We can then solve for $c_2$ like in \cref{linearly_dep_part_sol} to find the unique particular solution.

\end{document}