\documentclass[notes]{subfiles}

\begin{document}

\setcounter{section}{4}
\section{Video Lecture (1/22/25)}
\subsection{Linear Equations}

\begin{definition}[Linear First Order ODE]
    A first order ODE is \textit{linear} iff it can be written in the form
    \[
        a_1(x)y' + a_2(x)y = g(x)
    \]
    The \textit{standard form} for a linear first order ODE is
    \[
        y' + P(x)y = f(x)
    \]
\end{definition}
We call them linear differential equations because if you have $n$ solutions in a linear combination you have another solution.
Further reading: Linear differential operator

\begin{definition}[Integrating Factor]
    Given a first order linear ODE in the form $y' + P(x)y = f(x)$, then a function $\mu$ is an \textit{integrating factor} iff $\mu(x) = e^{\int P(x)dx}$
\end{definition}

We can solve any first order linear ODE in this standard form by multiplying through by an integrating factor.

\begin{theorem} \label{int_factor}
    Given a first order linear ODE in the form $y' + P(x)y = f(x)$ with an integrating factor $\mu$ then $y' + P(x)y = f(x) \iff y = \frac{1}{\mu(x)} \int \mu(x)f(x)dx$.
\end{theorem}
\begin{proof}
    \begin{align*}
        y' + P(x)y = f(x)
        &\iff \mu(x)(y' + P(x)y) = \mu(x)f(x) \\
        &\iff \mu(x)y' + e^{\int P(x)dx}P(x)y = \mu(x)f(x) \\
        &\iff \mu(x)y' + \mu'(x)y = \mu(x)f(x) \\
        &\iff \frac{d}{dx}(\mu(x)y) = \mu(x)f(x) \\
        &\iff \mu(x)y = \int \mu(x)f(x)dx \\
        &\iff y = \frac{1}{\mu(x)} \int \mu(x)f(x)dx
    \end{align*}
\end{proof}

\begin{procedure}[Solving first order linear ODEs] ~\par
    \begin{enumerate}[label = (\arabic*)]
        \item Ensure the equation isn't already separable.
        \item Write the equation in standard form.
        \item Find the integrating factor $\mu$.
        \item Multiply both sides of the equation by $\mu(x)$.
        \item Reinterpret with the product rule.
        \item Evaluate the resulting integral.
    \end{enumerate}
\end{procedure}

\begin{definition}[Transient Term]
    A term in a solution to a DE is \textit{transient} iff it goes to $0$ as $x \to \infty$ or $x \to -\infty$. We care about these terms since they often have a small contribution to the solution for large (in magnitude) values of $x$.
\end{definition}

\begin{example}
    Given the solution $y = c_1 e^{-x} + c_2 x^2$, then the term $c_1 e^{-x}$ is transient.
\end{example}

\begin{definition}[Singular Point]
    Given a first order linear ODE in the form $y' = P(x)y = f(x)$, then $x_0$ is a \textit{singular point} iff $P$ is undefined or discontinuous at $x_0$. In effect, $P$ is poorly behaved as $x \to x_0$
\end{definition}

\begin{exercise}
    What are the singular points of $x^2 y' + \cos(x)y = 0$?
\end{exercise}
\begin{solution}
    First we must get the equation in standard form so
    \begin{align*}
        x^2 y' + \cos(x)y = 0
        &\iff y' + \frac{\cos x}{x^2}y = 0
    \end{align*}
    So let $P(x) = \frac{\cos x}{x^2}$, notice that $0 \notin \dom P$ so $0$ is a singular point of the equation.
\end{solution}

\begin{exercise}
    Solve $xy' - 4y = x^6 e^x$.
\end{exercise}
\begin{solution}
    First, consider that
    \begin{align*}
        xy' - 4y = x^6 e^x
        &\iff xy' = x^6 e^x + 4y
        \iff y' = \frac{x^6 e^x + 4y}{x}
    \end{align*}
    But we cannot factor the RHS, so the equation is not separable. Now we need to put the equation in standard form so
    \begin{align*}
        xy' - 4y = x^6 e^x
        &\iff y' - 4\frac{1}{x}y = x^5 e^x
    \end{align*}
    So $P(x) = -4\frac{1}{x}$, now consider $\mu(x) = e^{\int P(x)dx}$ so
    \begin{align*}
        \mu(x)
        &= e^{\int P(x)dx}
        = e^{\int -4\frac{1}{x}dx}
        = e^{-4\ln|x|}
        = (e^{\ln|x|})^{-4}
        = x^{-4}
        = \frac{1}{x^4}
    \end{align*}
    Now we solve the equation so
    \begin{align*}
        y' - 4\frac{1}{x}y = x^5 e^x
        &\iff \frac{1}{x^4} y' - 4\frac{1}{x^5}y = x e^x
        \iff \frac{d}{dx}\left( \frac{1}{x^4}y \right) = x e^x \\
        &\iff \frac{1}{x^4}y = \int x e^x dx
    \end{align*}
    Now we need to evaluate the integral on the RHS so let $u = x \iff du = dx$ and $dv = e^x dx \iff v = e^x$
    \begin{align*}
        \int x e^x dx
        &= x e^x - \int e^x dx
        = C + x e^x - e^x
    \end{align*}
    Therefore
    \begin{align*}
        \frac{1}{x^4}y = \int x e^x dx
        &\iff \frac{1}{x^4}y = C + x e^x - e^x \\
        &\iff y = Cx^4 + x^5 e^x - x^4 e^x
    \end{align*}
    So $y = Cx^4 + x^5 e^x - x^4 e^x$ is our solution.
\end{solution}

\begin{exercise}
    Solve $y' = x + y$ where $y|_{x = 0} = 4$.
\end{exercise}
\begin{solution}
    This is trivially not separable. Now consider
    \begin{align*}
        y' = x + y
        &\iff y' - y = x
    \end{align*}
    So $\mu(x) = e^{\int -dx} = e^{-x}$. Therefore
    \begin{align*}
        y' - y = x
        &\iff e^{-x}y' - e^{-x}y = e^{-x}x \\
        &\iff \frac{d}{dx}(e^{-x}y) = e^{-x}x \\
        &\iff e^{-x}y = \int e^{-x}x dx
    \end{align*}
    For the integral let $u = x \iff du = dx$ and $dv = e^{-x}dx \iff v = -e^{-x}$
    \begin{align*}
        \int e^{-x}x dx
        &= -x e^{-x} + \int e^{-x}dx
        = C - x e^{-x} - e^{-x}
    \end{align*}
    Therefore
    \begin{align*}
        e^{-x}y = \int e^{-x}x dx
        &\iff e^{-x}y = C - x e^{-x} - e^{-x} \\
        &\iff y = Ce^x - x - 1
    \end{align*}
    Therefore $y = Ce^x - x - 1$ is our general solution. Now consider that
    \begin{align*}
        x = 0
        &\iff 4 = Ce^{0} - 0 - 1 = C - 1
        \iff C = 5
    \end{align*}
    Therefore $y = 5e^x - x - 1$ is our solution to the IVP. Furthermore, notice that $5e^x$ is a transient term so as $x \to -\infty$ we approach the line $y = -x - 1$.
\end{solution}

\end{document}